---
title: "BASS-GWAS Tutorial"
author: David Helekal
date: Aug 20, 2026
output: 
    md_document:
        toc: true
        toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=300,fig.width=7)
```

```{bash, warning=FALSE, message=FALSE, error=TRUE, eval=TRUE, include=FALSE}
rm -r round0
rm -r round1
rm -r round2
rm -r round3
rm -r round4
rm -r gwas_inputs
rm -r gwas_outputs
```

First we load the necessary libraries used in this tutorial
```{r, warning=FALSE, message=FALSE}
library(ape)
library(tidyverse)
library(ggplot2)

set.seed(1)
```

**NOTICE: RE-RUNNING THIS TUTORIAL REQUIRES ~16GB OF RAM AND 4 CPU CORES**

**EXPECTED EXECUTION TIME: ~25-45MIN**

# Introduction
This tutorial assumes that you have previously downloaded the files in this repository.
The first section of the tutorial covers the general usage of the GWAS model. 
The second section covers adaptive experimental design.

Throughout this tutorial we will be using the toy data found in the `sample_data` folder.
This is a version of the ciprofloxacin resistance benchmark dataset used in the manuscript modified to contain only 20,000 unique variants.
The toy dataset uses *unitigs* to characterise genetic variation.

# Input Data
Using BASS-GWAS requires two main data files: a variant presence-absence file, and a phylogeny file in the newick format. 
Both the phylogeny and variant presence-absence should be constructed for the entire collection at hand. The first column of the variant file should contain unique variant identifiers. 
The remaining columns should each correspond to an isolate, and the first row contain unique isolate identifiers:

|variant   |isolate_1|isolate_2|isolate_3| ... |isolate_n|
|----------|---------|---------|---------|-----|---------|
|variant_1 |1        |1        |0        | ... |1        |
|variant_2 |1        |1        |0        | ... |1        |
|variant_3 |0        |1        |1        | ... |0        |
|... |...        |...        |...        | ... |...      |
|variant_n |0        |0        |1        | ... |0        |

The isolate identifiers must be shared between the variant file and the phylogeny. 
If using unitigs, [unitig-caller](https://github.com/bacpop/unitig-caller) can create this file for you.

## Note on Input Types
Representing genetic variation in bacterial populations is an active area of research.
There are many distinct approaches, for example using VCFs, unitigs, kmers, gene presence-absence, etc.
To ensure proper functionality of BASS-GWAS, the representation of genetic variation should be both *comprehensive*, and *precise*.
This is especially important when using adaptive sequential sampling. Unitigs are usually a good choice. 

**DO NOT use adaptive sequential sampling with highly redundant representations such as kmers**

**DO NOT use adaptive sequential sampling with gene presence-absence data alone**

## Inspecting the input data
The variant file in the toy dataset uses unitigs as variant identifiers and `sra` accession numbers as isolate identifiers.
```{r show inputs, warning=FALSE}
vroom::vroom("../sample_data/variants.rtab",
    altrep=F, progress=F, show_col_types=F) %>%   
    print(n = 5, max_extra_cols = 0,width=60)
```
## Preparing the inputs
To generate the input files needed for use BASS-GWAS, we will need to use the data preprocessing script.
First we will create a directory for storing the generated input files.

```{bash, warning=FALSE, message=FALSE, error=TRUE, results = 'hide'}
mkdir gwas_inputs
```

Next we will use the data preprocessing script `prepare_data.R` to process the variant presence-absence file and the phylogeny into BASS-GWAS inputs.
As a part of preprocessing, variant presence-absence patterns will be standardised, centred, and deduplicated. The phylogeny is used to compute a factorised, low-rank similarity matrix.

`prepare_data.R` usage:
```{bash}
Rapp ../helper_scripts/prepare_data.R --help
```
- `--af-threshold` can be used to set an optional allele frequency filter. Variants with frequency lower than this value will be discarded. Default: 0.
- `--svd_threshold` controls the fraction of variance of the original phylogenetic correlation matrix that will be retained. Default: 0.99.

We will now use the script to preprocess the toy data. To speed up inference for the purposes of this tutorial, we will retain only 90% of the variance.
```{bash, warning=FALSE, message=FALSE}
Rapp ../helper_scripts/prepare_data.R ../sample_data/variants.rtab ../sample_data/tree.tre --odir gwas_inputs --svd_threshold 0.9
```
This will generate the following input files in a `gwas_inputs` subdirectory.

The following two files are subsequently used with BASS-GWAS:

- `patterns_centred.csv`: a centred, minorized, and deduplicated variant presence-absence file for use with BASS-GWAS.

- `U.csv` & `S.csv`: a low-rank factorisation of the correlation matrix implied by the phylogeny for use with BASS-GWAS.

The `vars2patts.csv` file maps between the original variant identifiers and the deduplicated variants that form `patterns_centred.csv`:

```{r show variant map,warning=FALSE}
    vroom::vroom("gwas_inputs/vars2patts.csv",
    altrep=F, progress=F, show_col_types=F) %>%   
    print(unitigs, n = 4, max_extra_cols = 0,width=60)
```
- `pattern_id` contains the column index in `patterns_centred.csv`. 

- `variant` contains the original variant identifier. 

- `direction` is equal to $$-1$$ if the original variant was complemented as $$1-v$$. 

The `patterns.csv` file is a `0/1` version of the pattern presence-absence file.

## Observations file
The observation file should be a table in the `.csv` format with two columns and a header.
The first column must contain strain identifiers. The second column must contain binary phenotype entries (either `0/1` or `TRUE/FALSE`).
An example is shown below:

|strain   |phenotype|
|----------|---------|
|isolate_1 |0   |
|isolate_2 |0   |
|isolate_3 |1   |
|isolate_4 |0   |
|isolate_10 |1   |
|... |...        |
|isolate_n |0  |

The toy dataset contains an observation file in the correct format:
```{r show observations, warning=FALSE}
vroom::vroom("../sample_data/data_full.csv",
    altrep=F, progress=F, show_col_types=F) %>%   
    print(n = 5, max_extra_cols = 0,width=60)
```

# GWAS
## `bassgwass-run.jl` usage
```
bassgwas-run.jl --pfile PFILE --ufile UFILE --sfile SFILE
                --obsfile OBSFILE --odir ODIR --batchsz BATCHSZ
                [--inclnext INCLNEXT] [--seed SEED]
                [--ndraws NDRAWS] [--nthin NTHIN]
                [--nchains NCHAINS] [--p_mean P_MEAN]
                [--pr_0 PR_0] [--randdes] [-h]
```

- `--pfile` `patterns_centred.csv` variant file (generated by preprocessing script)

- `--ufile` `U.csv` left singular vectors for population structure correction (generated by preprocessing script)

- `--sfile` `S.csv` singular values for population structure correction (generated by preprocessing script)

- `--obsfile` observations file

- `--odir` output directory

- `--batchsz` batch size for experimental design. If set to 0, experimental design will be disabled

- `--inclnext` path to a file listing isolate identifiers to include in the next experimental designs, one per line. If omitted, all isolates for next experimental design will be selected algorithmically. See [Additional functionality](#Additional-functionality)

- `--seed` random seed. Default: `0`

- `--n_draws` number of draws per chain. Default: `1000`

- `--n_thin` thinning factor per draw. Default: `500`

- `--n_chains` number of parallel chains to run. Default: `4`

- `--p_mean` prior mean number of causal variants conditional on at least one effect. Default: `5`

- `--pr_0` prior probability of 0 effects. Default: `0.25`

- `--randdes` flag indicated a random design should be returned rather than a design based on expected information gained. You probably don't want to use this.

- `-h --help` print help and exit. 

With the inputs ready, we can now run the GWAS model. 
BASS-GWAS uses efficient MCMC importance sampling and will create multiple files with MCMC draws when finished, together with a file specifying a batch of isolates for subsequent phenotyping if requested.

**NOTE the generated files are large. For 100k variants, 4000 draws require approximately 60GB of storage**

## Running GWAS on a toy dataset
We will now demonstrate GWAS functionality on the toy dataset. This dataset was constructed using data from (_Reimche 2023_).
The trait in this dataset is high-level ($$ \geq 1 \mu g \, mL^{-1}$$) ciprofloxacin resistance in _N. gonorrhoeae_.
To attain this level of resistance, at least one substitution in each of _gyrA_ and _parC_ genes is known to be required. 

The polygenic nature of this trait makes it challenging for GWAS with binary phenotypes.

We will run 2 chains in parallel each for a total of $$250 \times 500$$ iterations. 
Moreover we will parallelize each chain across 2 CPU cores using `julia -t2`. 
To speed things up, we will downsample the full dataset to 200 isolates.

```{r downsample data, warning=FALSE, message=FALSE}
dat_small <- read_csv("../sample_data/data_full.csv", progress=F, show_col_types=F) %>% 
    slice_sample(n = 200)
dat_small %>% group_by(cip) %>% 
    summarise(count=n())

dat_small %>% write_csv("gwas_inputs/data200.csv")
```

```{r, include=FALSE}
tbegin <- Sys.time()
```
We're now ready to run BASS-GWAS sampling. As the toy dataset only contains 20000 variants, this should only take a few minutes.
```{bash, warning=FALSE, message=FALSE, eval=TRUE}
mkdir gwas_outputs
julia -t2 ../bassgwas-run.jl --pfile gwas_inputs/patterns_centred.csv --ufile gwas_inputs/U.csv --sfile gwas_inputs/S.csv --obsfile gwas_inputs/data200.csv --odir gwas_outputs --batchsz 0 --ndraws 500 --nthin 250 --nchains 2
```

```{r}
paste0("Run time: ", difftime(Sys.time(),tbegin,units='mins'), " minutes.")
```

When using BASS-GWAS with real data, using up to 4 threads per chain is recommended. We recommend running chains for  $500\times 1000$ iterations and running 4 chains in parallel.

## Identifying associations 
### Background
BASS-GWAS uses a Bayesian sparse whole-genome regression model. The combined effect $$\psi$$ of all variants on the $j$-th strain is modelled as:

$$\psi_j = X_j\pmb{\beta} + u_j+\mu$$

Here $$\pmb{\beta}$$ is a vector of variant effects, $u_j$ is a random effect that accounts for population structure, and $\mu$ is an intercept. 
Bayesian sparse models ensure that all but a few of the entries of $$\pmb{\beta}$$ are zero, i.e. that all but a few variants have no effect. 

To identify association we will use *posterior inclusion probabilties* (PIPs). A PIP is the probability that a variant has a non-zero effect given the observed data. 

PIPs can be extended to *groups of variants*, for which the PIP is the probability that the specified group contains *at least* one variant with an effect given the observed data. 
This can be useful for example if we group variants by genomic position, or by gene annotation.

We can interpret PIPs as the probability of a variant, or a group of variants, containing a causal effect.

### Computing PIPs from BASS-GWAS outputs
BASS-GWAS saves the posterior draws into the `gwas_outputs` directory:
```{bash, warning=FALSE, message=FALSE}
ls gwas_outputs
```

- `b_draws.csv` variant coefficient draws

- `gamma_draws.csv` variant inclusion indicator draws

- `par_draws.csv` hyperparameter, summary statistic, and importance weight draws

- `pip_draws.csv` variant PIP draws

- `u_draws.csv` lineage effect draws

- `Y_draws.csv` latent utility draws for observed isolates

We will use the `comp_group_pips.R` script to identify associated loci using user-defined variant groups.

`comp_group_pips.R` usage:
```{bash}
Rapp ../helper_scripts/comp_group_pips.R --help
```

`comp_group_pips.R` computes PIPs for groups of variants specified in the variant group file.
The variant group file should list group-variant pairs:

|variant   |group|
|----------|---------|
|variant_1 |group1   |
|variant_1 |group2   |
|variant_2 |group4   |
|variant_3 |group2   |
|variant_3 |group3   |
|... |...        |
|variant_n |group_k  |

The toy dataset contains two variant grouping files:
 
 - `locus_groups.R` groups unitigs by annotations of the loci they map to.
 - `position_groups.R` groups unitigs by genomic position, using 250bp overlapping windows spaced 125bp apart.

```{r}
vroom::vroom("../sample_data/position_groups.csv",
    altrep=F, progress=F, show_col_types=F) %>%   
    print(n = 5, max_extra_cols = 0, width=60)
```

### Using Manhattan plots 

We can use grouping by genomic position, together with the posterior draws, to calculate the PIPs for individual genomic position windows along the chromosome.

```{bash, warning=FALSE, message=FALSE}
Rapp ../helper_scripts/comp_group_pips.R ../sample_data/position_groups.csv gwas_outputs/position_pips.csv gwas_outputs/par_draws.csv gwas_outputs/gamma_draws.csv gwas_inputs/vars2patts.csv
```

In turn, position PIPs can be visualised as a Manhattan plot:

```{r}
parC_begin <- 194768
parC_end <- 197071
parC_pos <- parC_begin+(parC_end-parC_begin)/2

gyrA_begin <- 996474
gyrA_end <- 999224
gyrA_pos <- gyrA_begin+(gyrA_end-gyrA_begin)/2
```

```{r, fig.dim = c(8, 2),dpi=300}
read_csv("gwas_outputs/position_pips.csv", progress=F, show_col_types=F) %>% 
    ggplot(aes(x=group/1e3, y=PIP)) + 
        geom_col(fill="gray25",color="gray25") +
        geom_vline(data=data.frame(x=c(parC_pos,gyrA_pos)/1e3), aes(xintercept=x), 
            color="darkred", linetype="dashed", alpha=.5, linewidth=1) +
        ylim(0,1)+
        labs(y="PIP", x="Position (KB)")
```

### Using variant annotations

Alternatively, we can use the locus grouping to identify associated annotated loci while limiting reference bias by mapping each unitig to the annotated de novo assembly from its own parent isolate.
```{bash, warning=FALSE, message=FALSE}
Rapp ../helper_scripts/comp_group_pips.R ../sample_data/locus_groups.csv gwas_outputs/locus_pips.csv gwas_outputs/par_draws.csv gwas_outputs/gamma_draws.csv gwas_inputs/vars2patts.csv
```

```{r}
read_csv("gwas_outputs/locus_pips.csv", progress=F, show_col_types=F) %>% 
    arrange(-PIP) %>% 
    print(n = 5, max_extra_cols = 0, width=60)
```

Based on these two summaries we can see that there are two peaks at the _gyrA_ and _parC_ loci, 
substitutions in both of which are required to attain high-level ciprofloxacin resistance. 

While _gyrA_ was identified with a relatively high level of confidence (PIP $$ > 0.7$$), 
the confidence for the _parC_ hit remains low (PIP $$ < 0.45$$). This is a consequence of random sampling.

Relying on hits with low PIPs can still lead to successful identification of causative variants but it will increase false positive rates, 
in turn leading to failed validation experiments. 

Moreover, with PIPs this low we cannot discern that __both__ _gyrA_ and _parC_ are required.

### Refining associated variants

We can further _fine-map_ the _gyrA_ hit to identify which variants are most likely to be causal. 
First we will compute the per-unitig PIPs.
```{bash, warning=FALSE, message=FALSE}
Rapp ../helper_scripts/comp_variant_pips.R gwas_outputs/variant_pips.csv gwas_outputs/par_draws.csv gwas_outputs/pip_draws.csv gwas_inputs/vars2patts.csv
```

Then we will use the variant annotation file to identify variants mapping to _gyrA_ and display 
the top 5.
```{r}
parC_hits <- read_csv("../sample_data/annotated_hits.csv", progress=F, show_col_types=F) %>% 
    filter(locus=="gyrA") %>%
    left_join(
        read_csv("gwas_outputs/variant_pips.csv", progress=F, show_col_types=F),
        by="variant")
        
parC_hits %>% 
    arrange(-PIP) %>% 
    print(n = 10, max_extra_cols = 0,width=60)
```

# Adaptive experimental design
As we saw above, even with 200 samples chosen at random, we were not able to identify _parC_ with reasonable confidence.
Instead, we will now conduct a synthetic experiment using the adaptive sampling functionality

## Setup
We will begin by subsetting the 200 isolates we used in the previous example to just 4, keeping at least one susceptible and one resistant isolate.
```{bash, warning=FALSE, message=FALSE, error=TRUE, results = 'hide'}
mkdir round0
```
```{r}
pheno_small <- dat_small$cip
idx1 <- sample(which(pheno_small), 1) 
idx0 <- sample(which(!pheno_small), 1) 
rem <- sample(c(1:length(pheno_small))[-c(idx1,idx0)], 4-2,replace=F)

to_keep <- c(idx0, idx1, rem)
write_csv(dat_small[to_keep,], "round0/data.csv")
```

## Adaptive design 
We will run BASS-GWAS adaptive design using a batch size of 16 for 4 rounds.

### Round 0
```{r, include=FALSE}
tbegin <- Sys.time()
```
Run BASS-GWAS
```{bash, warning=FALSE, message=FALSE, eval=TRUE}
julia -t2 ../bassgwas-run.jl --pfile gwas_inputs/patterns_centred.csv --ufile gwas_inputs/U.csv --sfile gwas_inputs/S.csv --obsfile round0/data.csv --odir round0 --batchsz 16 --ndraws 500 --nthin 250 --nchains 2
```

```{r}
paste0("Run time: ", difftime(Sys.time(),tbegin,units='mins'), " minutes.")
```
### Round 1
Create an output directory for round 1 and append the entries for the selected isolates
```{bash, warning=FALSE, message=FALSE, eval=TRUE}
mkdir round1
cp round0/data.csv round1/data.csv
grep -wf round0/design.txt ../sample_data/data_full.csv >> round1/data.csv
```
```{r, include=FALSE}
tbegin <- Sys.time()
```
Run BASS-GWAS
```{bash, warning=FALSE, message=FALSE, eval=TRUE}
julia -t2 ../bassgwas-run.jl --pfile gwas_inputs/patterns_centred.csv --ufile gwas_inputs/U.csv --sfile gwas_inputs/S.csv --obsfile round1/data.csv --odir round1 --batchsz 16 --ndraws 500 --nthin 250 --nchains 2
```

```{r}
paste0("Run time: ", difftime(Sys.time(),tbegin,units='mins'), " minutes.")
```
Compute position PIPs
```{bash, warning=FALSE, message=FALSE}
Rapp ../helper_scripts/comp_group_pips.R ../sample_data/position_groups.csv round1/position_pips.csv round1/par_draws.csv round1/gamma_draws.csv gwas_inputs/vars2patts.csv
```
Manhattan plot for round 1:
```{r, fig.dim = c(8, 2),dpi=300}
read_csv("round1/position_pips.csv", progress=F, show_col_types=F) %>% 
    ggplot(aes(x=group/1e3, y=PIP)) + 
        geom_col(fill="gray25",color="gray25") +
        geom_vline(data=data.frame(x=c(parC_pos,gyrA_pos)/1e3), aes(xintercept=x), 
            color="darkred", linetype="dashed", alpha=.5, linewidth=1) +
        ylim(0,1)+
        labs(y="PIP", x="Position (KB)")
```

### Round 2
Create an output directory for round 2 and append the entries for the selected isolates
```{bash, warning=FALSE, message=FALSE, class.source = 'fold-hide', eval=TRUE}
mkdir round2
cp round1/data.csv round2/data.csv
grep -wf round1/design.txt ../sample_data/data_full.csv >> round2/data.csv
```
```{r, include=FALSE}
tbegin <- Sys.time()
```
Run BASS-GWAS
```{bash, warning=FALSE, message=FALSE, class.source = 'fold-hide', eval=TRUE}
julia -t2 ../bassgwas-run.jl --pfile gwas_inputs/patterns_centred.csv --ufile gwas_inputs/U.csv --sfile gwas_inputs/S.csv --obsfile round2/data.csv --odir round2 --batchsz 16 --ndraws 500 --nthin 250 --nchains 2
```

```{r}
paste0("Run time: ", difftime(Sys.time(),tbegin,units='mins'), " minutes.")
```
Compute position PIPs
```{bash, warning=FALSE, class.source = 'fold-hide', message=FALSE}
Rapp ../helper_scripts/comp_group_pips.R ../sample_data/position_groups.csv round2/position_pips.csv round2/par_draws.csv round2/gamma_draws.csv gwas_inputs/vars2patts.csv
```
Manhattan plot for round 2:
```{r, fig.dim = c(8, 2),dpi=300}
read_csv("round2/position_pips.csv", progress=F, show_col_types=F) %>% 
    ggplot(aes(x=group/1e3, y=PIP)) + 
        geom_col(fill="gray25",color="gray25") +
        geom_vline(data=data.frame(x=c(parC_pos,gyrA_pos)/1e3), aes(xintercept=x), 
            color="darkred", linetype="dashed", alpha=.5, linewidth=1) +
        ylim(0,1)+
        labs(y="PIP", x="Position (KB)")
```

### Round 3
Create an output directory for round 3 and append the entries for the selected isolates
```{bash, warning=FALSE, message=FALSE, class.source = 'fold-hide', eval=TRUE}
mkdir round3
cp round2/data.csv round3/data.csv
grep -wf round2/design.txt ../sample_data/data_full.csv >> round3/data.csv
```
```{r, include=FALSE}
tbegin <- Sys.time()
```
Run BASS-GWAS
```{bash, warning=FALSE, message=FALSE, class.source = 'fold-hide', eval=TRUE}
julia -t2 ../bassgwas-run.jl --pfile gwas_inputs/patterns_centred.csv --ufile gwas_inputs/U.csv --sfile gwas_inputs/S.csv --obsfile round3/data.csv --odir round3 --batchsz 16 --ndraws 500 --nthin 250 --nchains 2
```

```{r}
paste0("Run time: ", difftime(Sys.time(),tbegin,units='mins'), " minutes.")
```
Compute position PIPs
```{bash, warning=FALSE, class.source = 'fold-hide', message=FALSE}
Rapp ../helper_scripts/comp_group_pips.R ../sample_data/position_groups.csv round3/position_pips.csv round3/par_draws.csv round3/gamma_draws.csv gwas_inputs/vars2patts.csv
```
Manhattan plot for round 3:
```{r, fig.dim = c(8, 2),dpi=300}
read_csv("round3/position_pips.csv", progress=F, show_col_types=F) %>% 
    ggplot(aes(x=group/1e3, y=PIP)) + 
        geom_col(fill="gray25",color="gray25") +
        geom_vline(data=data.frame(x=c(parC_pos,gyrA_pos)/1e3), aes(xintercept=x), 
            color="darkred", linetype="dashed", alpha=.5, linewidth=1) +
        ylim(0,1)+
        labs(y="PIP", x="Position (KB)")
```

### Round 4
Create an output directory for round 4 and append the entries for the selected isolates
```{bash, warning=FALSE, message=FALSE, class.source = 'fold-hide', eval=TRUE}
mkdir round4
cp round3/data.csv round4/data.csv
grep -wf round3/design.txt ../sample_data/data_full.csv >> round4/data.csv
```
```{r, include=FALSE}
tbegin <- Sys.time()
```
Run BASS-GWAS
```{bash, warning=FALSE, message=FALSE, class.source = 'fold-hide', eval=TRUE}
julia -t2 ../bassgwas-run.jl --pfile gwas_inputs/patterns_centred.csv --ufile gwas_inputs/U.csv --sfile gwas_inputs/S.csv --obsfile round4/data.csv --odir round4 --batchsz 16 --ndraws 500 --nthin 250 --nchains 2
```

```{r}
paste0("Run time: ", difftime(Sys.time(),tbegin,units='mins'), " minutes.")
```
Compute position PIPs
```{bash, warning=FALSE, class.source = 'fold-hide', class.source = 'foldable', message=FALSE}
Rapp ../helper_scripts/comp_group_pips.R ../sample_data/position_groups.csv round4/position_pips.csv round4/par_draws.csv round4/gamma_draws.csv gwas_inputs/vars2patts.csv
```
Manhattan plot for round 4:
```{r, fig.dim = c(8, 2),dpi=300}
read_csv("round4/position_pips.csv", progress=F, show_col_types=F) %>% 
    ggplot(aes(x=group/1e3, y=PIP)) + 
        geom_col(fill="gray25",color="gray25") +
        geom_vline(data=data.frame(x=c(parC_pos,gyrA_pos)/1e3), aes(xintercept=x), 
            color="darkred", linetype="dashed", alpha=.5, linewidth=1) +
        ylim(0,1)+
        labs(y="PIP", x="Position (KB)")
```
After round 4, representing a total of 68 (64 + 4) isolates both causal loci, parC and gyrA, were identified with high confidence (PIP $$ > 0.9$$). 
This makes it clear that variants in both loci contribute to high-level resistance. 
Further inspection of the predictor (beyond the scope of this tutorial) reveals the relationship between variants with high PIPs: in this case, that variants in both loci are required to achieve high-level resistance.

Let's see if the unitig-level PIPs are now more precise.
```{bash, warning=FALSE, message=FALSE}
Rapp ../helper_scripts/comp_variant_pips.R round4/variant_pips.csv round4/par_draws.csv round4/pip_draws.csv gwas_inputs/vars2patts.csv
```

Inspect _parC_ variants
```{r}
read_csv("../sample_data/annotated_hits.csv", progress=F, show_col_types=F) %>% 
    filter(locus=="parC" ) %>%
    left_join(
        read_csv("round4/variant_pips.csv", progress=F, show_col_types=F),
        by="variant")%>% 
    arrange(-PIP) %>% 
    print(n = 5, max_extra_cols = 0,width=60)
        
```
The variants mapping to _parC_ are dominated by a single unitig mapping to the QRDR region. 
This is in contrast to random sampling which failed to identify any _parC_ variants with high PIP.

Inspect _gyrA_ variants
```{r}
read_csv("../sample_data/annotated_hits.csv", progress=F, show_col_types=F) %>% 
    filter(locus=="gyrA" ) %>%
    left_join(
        read_csv("round4/variant_pips.csv", progress=F, show_col_types=F),
        by="variant")%>% 
    arrange(-PIP) %>% 
    print(n = 5, max_extra_cols = 0,width=60)
        
```
The variants mapping to _gyrA_ are dominated by a handful of strongly associated variants all mapping to the QRDR region.
The relatively low PIPs for each of these individual variants reflects the unitig representation for this highly diverse causal locus within _gyrA_, which is split over several different unitigs.

In both cases adaptive design greatly improved efficiency over random sampling.

## Additional functionality
BASS-GWAS provides additional functionality that may be useful in real-world experimental applications.

### Re-using draws to redesign an experiment
`bassgwas-design.jl` allows for experimental design to be optimized using a previous MCMC run.

```
bassgwas-design.jl --pfile PFILE --ufile UFILE --sfile SFILE
                        --obsfile OBSFILE --b_draws B_DRAWS
                        --u_draws U_DRAWS --par_draws PAR_DRAWS
                        --ofile OFILE --batchsz BATCHSZ
                        [--inclnext INCLNEXT]
```
- `--pfile` `patterns_centred.csv` variant file

- `--ufile` `U.csv` left singular vectors for population structure correction

- `--sfile` `S.csv` singular values for population structure correction

- `--obsfile` observations file

- `--b_draws` `b_draws.csv` variant coefficient draws

- `--u_draws` `u_draws.csv` lineage effect draws

- `--par_draws` `par_draws.csv` hyperparameter, summary statistic, and importance weight draws

- `--ofile` output file name

- `--batchsz` batch size for experimental design. If set to 0, experimental design will be disabled

- `--inclnext` path to a file listing isolate identifiers to include in the next experimental designs, one per line

- `-h --help` print help and exit. 

### Including fixed isolates in design 
The `--inclnext` argument can be uses to pass a text file listing isolate identifiers to include in the design. One per line.

### Removing problematic isolates from the collection
During experimental design, problematic or non-viable isolates may be encountered.
To drop problematic isolates, we can generate new BASS-GWAS inputs using the original variant and tree file, and pass a list of isolate identifiers to exclude to `prepare_data.R`, one identifier per line.

```
prepare_data.R --to_drop identifiers_to_drop.txt
```

# Tips
- Ensure that only high-quality isolates are included in experimental design collections. Mislabeled or contaminated isolates may severely compromise BASS-GWAS functionality.

- Use experimental controls to reduce phenotyping noise. 

- Set the allele frequency filter to a reasonable value. This should be organism specific. A frequency of $$5/N$$ or $$10/N$$ where N is the collection size is a good starting value.

- When applying BASS-GWAS to a new collection or a new organism, run synthetic experiments using known genotype/phenotype relationships to ensure proper data preparation.

- Check the tempered distribution diagnostics. The `ess_bulk` should ideally be more than 1000 for each of the parameters, and more than 200 per chain. If the `ess_bulk` is low, increase `--n_thin` to a higher value.
