---
title: "BASSGWAS Tutorial"
author: David Helekal
date: Aug 20, 2026
output: 
    md_document:
        toc: true
---

First we load the necessary libraries used in this tutorial
```{r setup, warning=FALSE, message=FALSE}
library(ape)
library(ggtree)
library(tidyverse)
library(ggplot2)

set.seed(1)
```

```{bash, warning=FALSE, message=FALSE, error=TRUE}
rm -r round1
rm -r round2
rm -r round3
rm -r gwas_inputs
rm -r gwas_outputs
```

# Introduction
This tutorial assumes that you have previously downloaded the helper & launch scripts in this repository.
The first section of the tutorial covers the general usage of the GWAS model. 
The second section covers adaptive experimental design.

Throughout this tutorial we will be using the toy data found in the `sample_data` folder.
This is a version of the ciprofloxacin resistance benchmark datasets used in the manuscript modified to contain only 20,000 unique variants.
The toy dataset uses *unitigs* to characterise genetic variation.

# Input Data
Using BASSGWAS requires two main data files: a variant presence-absence file, and a phylogeny file in the newick formati. 
Both the phylogeny and variant presence-absence should be constructed for the entire collection at hand. The first column of the variant file should contain unique variant identifiers. 
The remaining columns should each correspond to an isolate, and the first row contain unique isolate identifiers:

|variant_id|isolate_1|isolate_2|isolate_3| ... |isolate_n|
|----------|---------|---------|---------|-----|---------|
|variant_1 |1        |1        |0        | ... |1        |
|variant_2 |1        |1        |0        | ... |1        |
|variant_3 |0        |1        |1        | ... |0        |
|... |...        |...        |...        | ... |...      |
|variant_n |0        |0        |1        | ... |0        |

The isolate indentifiers must be shared between the variant file and the phylogeny. 



## Note on Input Types
Representing genetic variation in bacterial populations is an active area of research.
There are many distinct approaches, for example using VCFs, unitigs, kmers, gene presence absence, etc.
To ensure proper functionality of BASSGWAS, the representation of genetic variation should be both *comprehensive*, and *precise*.
This is especially important when using adaptive sequential sampling. Unitigs are usually a good choice. 

**DO NOT use adaptive sequential sampling with highly redundant representations such as kmers**

**DO NOT use adaptive sequential sampling with gene presence absence data alone**

## Inspecting the input data
The variant file in the toy dataset uses unitigs as variant identifiers and `sra` accession numbers as isolate identifiers.
```{r show inputs,warning=FALSE}
vroom::vroom("../sample_data/variants.rtab",
    altrep=F, progress=F, show_col_types = FALSE) %>%   
    print(unitigs, n = 5, max_extra_cols = 0,width=60)
```
## Preparing the inputs
To generate the input files needed for use BASSGWAS, we will need to use the data preprocessing script.
First we will create a directory for storing the generated input files.

```{bash, warning=FALSE, message=FALSE, results = 'hide'}
mkdir gwas_inputs
```

Next we will use the data preprocessing script `prepare_data.R` to process the variant presence-absence file and the phylogeny into BASSGWAS inputs.
As a part of preprocessing, variant presence absence patterns will be standardised, centred, and deduplicated. The phylogeny is used to compute a factorised, low rank, similarity matrix.
```{bash}
Rapp ../helper_scripts/prepare_data.R --help
```
The `--af-threshold` can be used to set an optional allele frequency filter. 

We will now use the script to preprocess the toy data.
```{bash, warning=FALSE, message=FALSE, results = 'hide'}
Rapp ../helper_scripts/prepare_data.R ../sample_data/variants.rtab ../sample_data/tree.tre --odir gwas_inputs
```
This generated input files in the `gwas_inputs` directory. 
The following two files are subsequently used with BASSGWASS:

- `patterns_centred.csv` a centred, minorized, and deduplicated variant presence absence file for use with BASSGWAS.

- `U.csv` & `S.csv` a low factorisation of the correlation matrix implied by the phylogeny for use with BASSGWAS.
 
The `vars2patts.csv` file maps between the original variant identifiers and the deduplicated variants that form `patterns_centred.csv`:
```{r show variant map,warning=FALSE}
vroom::vroom("gwas_inputs/vars2patts.csv",
    altrep=F, progress=F, show_col_types = FALSE) %>%   
    print(unitigs, n = 4, max_extra_cols = 0,width=60)
```

- `pattern_id` contains the column index in `patterns_centred.csv`. 

- `variant` contains the original variant identifier. 

- `direction` is equal to $-1$ if the original variant was complemented as $1-v$. 

Additionally an uncentred, binary version of the pattern file was generated as `patterns.csv`.

## GWAS
### `bassgwass-run.jl` usage
```
bassgwas-run.jl --pfile PFILE --ufile UFILE --sfile SFILE
                --obsfile OBSFILE --odir ODIR --batchsz BATCHSZ
                [--inclnext INCLNEXT] [--seed SEED]
                [--ndraws NDRAWS] [--nthin NTHIN]
                [--nchains NCHAINS] [--p_mean P_MEAN]
                [--pr_0 PR_0] [--randdes] [-h]
```
###

- `--pfile` `patterns_centred.csv` variant file

- `--ufile` `U.csv` left singular vectors for population structure correction

- `--sfile` `S.csv` singular values for population structure correction

- `--odir` output directory

- `--batchsz` batch size for experimental design. If set to 0, experimental design will be disabled

- `--inclnext` path to a file listing isolate identifiers to include in the next experimental designs, one per line

- `--seed` random seed. Default: `0`

- `--n_draws` number of draws per chain. Default: `1000`

- `--n_thin` thinning factor per draw. Default: `500`

- `--n_chains` number of parallel chains to run. Default: `4`

- `--p_mean` prior mean number of causal variants conditional on at least one effect. Default: `5`

- `--pr_0` prior probability of 0 effects. Default: `0.25`

- `--randdes` flag indicated a random design should be returned. You probably don't want to use this.

- `-h --help` print help and exit. 

With the inputs ready, we can now run the GWAS model. 
BASSGWAS uses efficient MCMC sampling and will create multiple files with MCMC draws when finished, 
along with a file specifying a batch of isolates for subsequent phenotyping if requested.

**NOTE the generated files are large. For 100k variants, 4000 draws require approximately 60GB of storage**

### Running GWAS on the toy dataset
We will now demonstrate GWAS functionality on the toy dataset. We will run 2 chains in parallel each for a total of $250\times 500$ iterations. Moreover we will parallelize each chain across 2 CPU cores using `julia -t2`. To speed things up, we will downsample the full dataset to 200 isolates.

```{r downsample data, warning=FALSE, message=FALSE}
dat_small <- read_csv("../sample_data/data_bin.csv") %>% 
    slice_sample(n = 200)
dat_small %>% group_by(cip) %>% 
    summarise(count=n())

dat_small %>% write_csv("gwas_inputs/data200.csv")
```

We're now ready to run BASSGWAS. As the toy dataset only contains 10000 variants, this should only take a few minutes.
```{bash, warning=FALSE, message=FALSE}
mkdir gwas_outputs
julia -t2 ../bassgwas-run.jl --pfile gwas_inputs/patterns_centred.csv --ufile gwas_inputs/U.csv --sfile gwas_inputs/S.csv --obsfile gwas_inputs/data200.csv --odir gwas_outputs --batchsz 0 --ndraws 500 --nthin 250 --nchains 2
```

When using BASSGWAS with real data using up to 4 threads per chain is recommended. It's a good idea to aim for at least 250 samples per chain.

### Identifying associations 

When using BASSGWAS with real data using up to 4 threads per chain is recommended.

## Adaptive experimental design




